---
title: "R Notebook"
output: html_notebook
---


```{r setup}
library(rstan)
rstan_options(auto_write = TRUE)
#options(mc.cores = parallel::detectCores())
options(mc.cores = 3)
library(tidyverse)
library(loo)
library(MVN)
library(bayesplot)
```


# Load the sigB data from GSE6865

The original source of data: (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE6865)

```{r}
temp <- tempfile();
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE6nnn/GSE6865/matrix/GSE6865_series_matrix.txt.gz", temp)
gse6865_raw_df = read.delim(gzfile(temp), comment.char = "!") #Intermediate data frame representation

# Raw profile data. We scale by 1000 to get to more wieldy values
gse6865_raw = as.matrix(gse6865_raw_df[,2:15]) / 1000
rownames(gse6865_raw) = gse6865_raw_df$ID_REF

#Times (in minutes) for the individual samples
gse6865_raw_time = c(0,5,10,15,20,25,30,40,50,60,70,80,90,100)
colnames(gse6865_raw) <- sapply(gse6865_raw_time,  FUN = function(x) { paste0(x,"min")})

# For the first measurement, we can expect the value to be 0 (the series are from germination)
gse6865_raw[is.na(gse6865_raw[,1]),1] = 0

# For the second measurement we will go with a linear interpolation
na2 = is.na(gse6865_raw[,2])
gse6865_raw[na2,2] = 0.5 * gse6865_raw[na2,1] + 0.5 * gse6865_raw[na2,3]

#cleanup the intermediate results
rm(gse6865_raw_df)
rm(na2)
rm(temp)

```
```{r}
measurement_times = floor(gse6865_raw_time / 2) + 1
smooth_time  <- seq(0,100, by = 2)
expression_data <- gse6865_raw

spline_df <- 6
spline_basis <- bs(smooth_time, degree = 3, df = spline_df)
sigB_spline_params <- spline_params(
  spline_basis = spline_basis,
  scale = 5
)

sigB_params_prior <-  params_prior(
  initial_condition_prior_sigma = 2,
  asymptotic_normalized_state_prior_sigma = 0.5,
  degradation_prior_mean = -2,
  degradation_prior_sigma = 1,
  mean_regulatory_input_prior_sigma = 5,
  sd_regulatory_input_prior_sigma =5,
  intercept_prior_sigma = 2
)

```

#Estimate sigma_absolute
```{r}
sigma_absolute_model <- stan_model(file = 'estimate_sigma_absolute.stan')
```

```{r}
low_expression_genes <- c("abfA","dppD", "feuC","ytmI","yqdB","ykrU")
data_sigma_absolute = list(
  num_measurements = length(measurement_times),
  num_targets = length(low_expression_genes),
  expression = expression_data[low_expression_genes,],
  sigma_absolute_prior_sigma = 0.2,
  sigma_relative = 0.3
)

fit_sigma_absolute <- sampling(sigma_absolute_model, data = data_sigma_absolute)
fit_sigma_absolute
sigma_absolute_samples <- rstan::extract(fit_sigma_absolute)$sigma_absolute
sigma_absolute_mean <- mean(sigma_absolute_samples)
sigma_absolute_sigma <- sd(sigma_absolute_samples)
density_points <- seq(min(sigma_absolute_samples), max(sigma_absolute_samples), length.out = 100)
ggplot(data.frame(x = sigma_absolute_samples), aes(x=x)) + geom_density() + 
  geom_line(data = data.frame(x=density_points, y = dtruncnorm(density_points, a = 0, mean = sigma_absolute_mean, sd = sigma_absolute_sigma)), aes(y = y), color = "blue")
```


#Multiple targets model
```{r}
multiple_targets_model <- stan_model(file = 'multiple_targets_splines.stan')
```

#Estimate sigma with splines
```{r}

sources <- c("sigA","sigB", sample(rownames(expression_data), 5))

data_sigma_estimate <- regulation_model_params(
  measurement_times = measurement_times,
  regulator_expression = t(expression_data[sources,]),
  measurement_sigma = measurement_sigma_prior(
      sigma_absolute_prior_mean = 0,
      sigma_relative_prior_mean = 0,
      sigma_absolute_prior_sigma = 0.1,
      sigma_relative_prior_sigma = 0.2
  ),
  params_prior = sigB_params_prior,
  spline_params = sigB_spline_params
)

fit_sigma_estimate <- sampling(multiple_targets_model, data = data_sigma_estimate, control = list(adapt_delta = 0.95))

summary(fit_sigma_estimate, pars = c("sigma_relative_param","sigma_absolute_param"))$summary

for(s in 1:length(sources)) {
  print(fitted_regulator_plot(fit_sigma_estimate, data_sigma_estimate, s))
}
```


#With known regulator

##Show splining if no target info was used
```{r}
source <- "sigA"
#source <- "sigB"
targets <- c()

data <- regulation_model_params(
  measurement_times = measurement_times,
  regulator_expression = expression_data[source,],
  # measurement_sigma = measurement_sigma_prior(
  #     sigma_absolute_prior_mean = 0,
  #     sigma_relative_prior_mean = 0,
  #     sigma_absolute_prior_sigma = 0.1,
  #     sigma_relative_prior_sigma = 0.2
  # ),
  measurement_sigma = measurement_sigma_given(0.1,0.2),
  spline_params = sigB_spline_params,
  params_prior = sigB_params_prior
)


fit <- sampling(multiple_targets_model, data = data, control = list(adapt_delta = 0.95))
print(summary(fit, pars = c("coeffs", "intercept","initial_condition","sensitivity","degradation","w","b"))$summary)

fitted_regulator_plot(fit, data)
```

# Improved splining where target info is use


```{r}
source <- "sigA"
#source <- "sigB"
targets <- c("codY","clpQ","yqxD","yhdL", "comA","codV","comEB", "dnaG")#, "ftsA","hemD","acsA","acuB")
#targets <- c("codY","clpQ","yqxD")


data_source <- regulation_model_params(
  measurement_times = measurement_times,
  target_expression = t(expression_data[targets,]),
  regulator_expression = expression_data[source,],
  # measurement_sigma = measurement_sigma_prior(
  #     sigma_absolute_prior_mean = 0,
  #     sigma_relative_prior_mean = 0,
  #     sigma_absolute_prior_sigma = 0.1,
  #     sigma_relative_prior_sigma = 0.2
  # ),
  measurement_sigma = measurement_sigma_given(0.2,0.05),
  spline_params = sigB_spline_params,
  params_prior = sigB_params_prior,
  regulation_signs = matrix(1, 1, length(targets))
)

fit_source <- sampling(multiple_targets_model, data = data_source, control = list(adapt_delta = 0.95))

#print(summary(fit_source, pars = c("sensitivity","degradation","w","b","sigma_absolute_param","sigma_relative_param"))$summary)

num_samples <- 100
main_geom <- geom_line(alpha = 0.2)
total_samples <- length(rstan::extract(fit_source,"sigma_relative")$sigma_relative)
samples_to_show <-  sample(1:total_samples, num_samples)
samples_regulator <- rstan::extract(fit_source,"predicted_regulator_expression")$predicted_regulator_expression[samples_to_show,,1]

regulator_plot <- ggmatplot(1:data_source$num_time, t(samples_regulator), main_geom = main_geom) + ggtitle(paste0("Regulator - ", source))

if(data_source$regulators_measured != 0) {
  regulator_plot <- regulator_plot + geom_point(data = data.frame(x = data_source$measurement_times,  y = data_source$regulator_expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)
}
print(regulator_plot)
 
samples_expression_replicates <- rstan::extract(fit_source,"expression_replicates")$expression_replicates[samples_to_show,,,drop=FALSE]
for(target in 1:length(targets)) {

  
  print(ggmatplot(data_source$measurement_times, t(samples_expression_replicates[,,target]), main_geom = main_geom, x_title = "time", y_title = "expression") +
    geom_point(data = data.frame(x = data_source$measurement_times,  y = data_source$expression[,target]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Replicates - ", targets[target])))
  
}
```


##Checking whether resulting coeffs are approximately multivariate normal

```{r}
pairs(fit_source, pars = "coeffs")
samples_coeffs <- rstan::extract(fit_source,"coeffs")$coeffs[,,1]
uniPlot(samples_coeffs, type ="qqplot")
uniPlot(samples_coeffs, type = "histogram")
mardiaTest(samples_coeffs, qqplot = TRUE)
mvOutlier(samples_coeffs, qqplot = TRUE, method = "adj.quan")
```
We see there are some deviations from normality. In particular it seems the posterior has heavier tails than normal. Let's see what happens when we try to fit multivariete normal anyway. 

```{r}
coeffs_prior_cov <- 0.5 * cov(samples_coeffs)
chol_cov <- chol(coeffs_prior_cov)
coeffs_prior_mean <- colMeans(samples_coeffs)

means_array <- t(array(
  rep(coeffs_prior_mean,dim(samples_coeffs)[1]),c(dim(samples_coeffs)[2],dim(samples_coeffs)[1])))

samples_coeffs_mvn <- array(rnorm(dim(samples_coeffs)[1] * dim(samples_coeffs)[2], 0, 1), dim(samples_coeffs)) %*% chol_cov +  means_array

measured_geom <- geom_point(data = data.frame(x = data_source$measurement_times,  y = data_source$regulator_expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)

num_samples = 100
samples_to_show <- sample(1:nrow(samples_coeffs_mvn),num_samples)
spline_samples <- t(samples_coeffs_mvn[samples_to_show,] %*% t(data_source$spline_basis) ) * data_source$scale
ggmatplot(1:data_source$num_time, spline_samples, main_geom = main_geom) + ggtitle("Regulator - MVN") + measured_geom

spline_samples_orig <- t(samples_coeffs[samples_to_show,] %*% t(data_source$spline_basis) ) * data_source$scale
ggmatplot(1:data_source$num_time, spline_samples_orig, main_geom = main_geom) + ggtitle("Regulator - Orig") + measured_geom


```


#Using the fitted normal to check regulations of other genes

```{r}
set.seed(132468714)
targets <- sample(rownames(expression_data[])[rowMeans(expression_data) > 1])
```



```{r}
#targets <- c("yfkJ")
#targets <- c("yqxD","yfkJ","pyrB","purT")
#targets <- c("yqxD")
#targets <- c("jag","kinC","kinE","lepA")
targets <- c("codV", "yhdL", "kinE","purT")
#tagG, rplD, ptsG, ahpC, glmS, yurV

data_target <- list()
for(t in 1:length(targets)) {

 data_target[[t]] <- regulation_model_params(
    measurement_times = measurement_times,
    target_expression = expression_data[targets[t],],
    #regulator_expression = expression_data[source,],
    coeffs_prior = coeffs_prior_from_fit(fit_source, covariance_scale = 1),
    # measurement_sigma = measurement_sigma_prior(
    #     sigma_absolute_prior_mean = 0,
    #     sigma_relative_prior_mean = 0,
    #     sigma_absolute_prior_sigma = 0.1,
    #     sigma_relative_prior_sigma = 0.2
    # ),
    measurement_sigma = measurement_sigma_given(0.2,0.05),
    spline_params = sigB_spline_params,
    params_prior = sigB_params_prior,
    regulation_signs = matrix(1, 1, 1)
  )
 
}

results_target <- fit_multiple_targets_multi(data_target, output.dir = "out/test_target")

fits_target <- list()
   
for(t in 1:length(targets)) {
  fits_target[[t]] <- sampling_multi_read_fit(results_target, t)
  #print(fitted_regulator_plot(fits_target[[t]], data_target[[t]])) 
  print(fitted_target_plot(fits_target[[t]],data_target[[t]], name = targets[t]))
  print(fitted_target_observed_plot(fits_target[[t]],data_target[[t]], name = targets[t]))
}

```


TODO: zkorelovat WAIC vs. sum sq.

```{r}

looic <- array(-Inf,length(targets))
waic <- array(-Inf,length(targets))
for(t in 1:length(targets)) {
  if(t > length(fits_target) || is.null(fits_target[[t]])) {
    next
  }
  samples_log_lik <- rstan::extract(fits_target[[t]], "log_likelihood")$log_likelihood
  lik_data <- samples_log_lik[,,1]
  loo_res <- loo(lik_data)
  looic[t] <- loo_res$looic
  waic[t] <- waic(lik_data)$waic
}
o <- order(looic)
data.frame(order = (1:length(targets))[o],gene = targets[o], waic = waic[o], looic = looic[o]) %>% filter(is.finite(looic))

```

##Comparing to constant synthesis
```{r}
csynth_model <- stan_model(file = 'constant_synthesis.stan')
```

```{r}
fits_csynth <- list()
for(t in 1:length(targets)) {
  if(t > length(fits_target) || is.null(fits_target[[t]])) {
    next
  }
  
  data <- list(
    num_measurements = length(measurement_times),
    measurement_times = measurement_times,
    expression = expression_data[targets[t],],
    measurement_sigma_absolute = 0.2,
    measurement_sigma_relative = 0.05,
    initial_condition_prior_sigma = 2,
    asymptotic_normalized_state_prior_sigma = 0.5,
    degradation_prior_mean = -2,
    degradation_prior_sigma = 1
  )  
  fit_csynth <- sampling(csynth_model, data = data, control = list(adapt_delta = 0.95))
   summary(fit_csynth)$summary[,c("n_eff","Rhat")] %>% as.data.frame() %>% summarise(min_n_eff = min(n_eff),max_Rhat = max(Rhat))

  num_samples <- 100
  main_geom <- geom_line(alpha = 0.2)
  samples_to_show <-  sample(1:4000, num_samples)
  
  
  samples_expression <- rstan::extract(fit_csynth,"predicted_expression")$predicted_expression[samples_to_show,]
  print(ggmatplot(data$measurement_times, t(samples_expression), main_geom = main_geom) + 
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(targets[t]))
  
  samples_replicates <- rstan::extract(fit_csynth,"expression_replicates")$expression_replicates[samples_to_show,]
  print(ggmatplot(data$measurement_times, t(samples_replicates), main_geom = main_geom) +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(targets[t]))
  
  fits_csynth[[t]] <- fit_csynth
}
```
```{r}

looic <- array(-1,length(targets))
waic <- array(-1,length(targets))
for(t in 1:length(targets)) {
  if(t > length(fits_csynth) || is.null(fits_csynth[[t]])) {
    next
  }

  samples_log_lik <- rstan::extract(fits_csynth[[t]], "log_likelihood")$log_likelihood
  lik_data <- samples_log_lik
  loo_res <- loo(lik_data)
  looic[t] <- loo_res$looic
  waic[t] <- waic(lik_data)$waic
}
o <- order(looic)
data.frame(order = (1:length(targets))[o],gene = targets[o], waic = waic[o], looic = looic[o]) %>% filter(waic >= 0)

```


##Without known regulator

```{r}
source <- "sigA"
targets <- c("codY","clpQ","yqxD") #,"dnaA","yfkJ","yhdL", "pyrB","purT"

data <- list(
  num_time = ceiling(length(smooth_time) / 2),
  num_measurements = length(gse6865_raw_time),
  num_targets = length(targets),
  regulator_measured = 0,
  measurement_times = floor(gse6865_raw_time / 2) + 1,
  num_knots = num_knots,
  knots = knots / 2,
  spline_degree = 3,
  expression = gse6865_raw[targets,],
  regulator_expression = numeric(0),#gse6865_raw[source,],
  regulation_signs = array(1, length(targets)),
  measurement_sigma_absolute = 0.2,
  measurement_sigma_relative = 0.1,
  intercept_prior_sigma = numeric(0),
  initial_condition_prior_sigma = 2,
  mean_synthesis_prior_sigma = 1,
  degradation_prior_mean = -3,
  degradation_prior_sigma = 1,
  w_prior_sigma = 2,
  b_prior_sigma = 5,
  scale = 5
)

fit <- sampling(multiple_targets_model, data = data, control = list(adapt_delta = 0.95))
print(summary(fit, pars = c("coeffs", "intercept","initial_condition","sensitivity","degradation","w","b"))$summary)

num_samples <- 100
main_geom <- geom_line(alpha = 0.2)
samples_to_show <-  sample(1:4000, num_samples)
samples_regulator <- rstan::extract(fit,"predicted_regulator_expression")$predicted_regulator_expression[samples_to_show,]

regulator_plot <- ggmatplot(1:data$num_time, t(samples_regulator), main_geom = main_geom) + ggtitle(paste0("Regulator - ", source))

if(data$regulator_measured != 0) {
  regulator_plot <- regulator_plot + geom_point(data = data.frame(x = data$measurement_times,  y = data$regulator_expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)
}
print(regulator_plot)
 
samples_expression <- rstan::extract(fit,"predicted_expression")$predicted_expression[samples_to_show,,,drop=FALSE]
for(target in 1:length(targets)) {
  print(ggmatplot(1:data$num_time, t(samples_expression[,target,]), main_geom = main_geom) +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression[target,]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Expression - ", targets[target])))
}
  

```
