---
title: "R Notebook"
output: html_notebook
---


```{r setup}
library(rstan)
rstan_options(auto_write = TRUE)
#options(mc.cores = parallel::detectCores())
options(mc.cores = 3)
library(tidyverse)
library(loo)
library(MVN)
```


# Load the sigB data from GSE6865

The original source of data: (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE6865)

```{r}
temp <- tempfile();
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE6nnn/GSE6865/matrix/GSE6865_series_matrix.txt.gz", temp)
gse6865_raw_df = read.delim(gzfile(temp), comment.char = "!") #Intermediate data frame representation

# Raw profile data. We scale by 1000 to get to more wieldy values
gse6865_raw = as.matrix(gse6865_raw_df[,2:15]) / 1000
rownames(gse6865_raw) = gse6865_raw_df$ID_REF

#Times (in minutes) for the individual samples
gse6865_raw_time = c(0,5,10,15,20,25,30,40,50,60,70,80,90,100)
colnames(gse6865_raw) <- sapply(gse6865_raw_time,  FUN = function(x) { paste0(x,"min")})

# For the first measurement, we can expect the value to be 0 (the series are from germination)
gse6865_raw[is.na(gse6865_raw[,1]),1] = 0

# For the second measurement we will go with a linear interpolation
na2 = is.na(gse6865_raw[,2])
gse6865_raw[na2,2] = 0.5 * gse6865_raw[na2,1] + 0.5 * gse6865_raw[na2,3]

#cleanup the intermediate results
rm(gse6865_raw_df)
rm(na2)
rm(temp)

```
```{r}
measurement_times = floor(gse6865_raw_time / 2) + 1
smooth_time  <- seq(0,100, by = 2)
spline_df <- 6
spline_basis <- t(bs(smooth_time, degree = 3, df = spline_df))
num_spline_basis <- dim(spline_basis)[1]
gse6865_splined <- spline_profile_matrix(gse6865_raw, gse6865_raw_time, smooth_time, df = spline_df)
```

#Multiple targets model
```{r}
multiple_targets_model <- stan_model(file = 'multiple_targets_splines.stan')
```

#With known regulator

##Show splining if no target info was used
```{r}
source <- "sigA"
#source <- "sigB"
targets <- c()

data <- list(
  num_time = length(smooth_time),
  num_measurements = length(gse6865_raw_time),
  num_targets = length(targets),
  regulator_measured = 1,
  
  coeffs_prior_given = 0, #No coeffs prior. Sadly have to give at least size 1
  coeffs_prior_mean = array(0.0,1),
  coeffs_prior_cov = array(1, c(1,1)),
  
  measurement_times = measurement_times,
  num_spline_basis = num_spline_basis,
  spline_basis = spline_basis,
  expression = gse6865_raw[targets,],
  regulator_expression = gse6865_raw[source,],
  regulation_signs = array(1, length(targets)),
  measurement_sigma_absolute = 0.2,
  measurement_sigma_relative = 0.1,
  intercept_prior_sigma = 2,
  initial_condition_prior_sigma = 0.05,
  sensitivity_over_degradation_prior_mean = 2.5,
  sensitivity_over_degradation_prior_sigma = 2,
  sensitivity_prior_sigma = 1,
  degradation_prior_mean = -3,
  degradation_prior_sigma = 1,
  w_prior_sigma = 2,
  b_prior_sigma = 5,
  scale = 5
)

fit <- sampling(multiple_targets_model, data = data, control = list(adapt_delta = 0.95))
print(summary(fit, pars = c("coeffs", "intercept","initial_condition","sensitivity","degradation","w","b"))$summary)

num_samples <- 100
main_geom <- geom_line(alpha = 0.2)
samples_to_show <-  sample(1:4000, num_samples)
samples_regulator <- rstan::extract(fit,"predicted_regulator_expression")$predicted_regulator_expression[samples_to_show,]

regulator_plot <- ggmatplot(1:data$num_time, t(samples_regulator), main_geom = main_geom) + ggtitle(paste0("Regulator - ", source))

if(data$regulator_measured != 0) {
  regulator_plot <- regulator_plot + geom_point(data = data.frame(x = data$measurement_times,  y = data$regulator_expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)
}
print(regulator_plot)
```

# Improved splining where target info is use

```{r}
source <- "sigA"
#source <- "sigB"
targets <- c("codY","clpQ","yqxD","yhdL", "comA","codV","comEB", "dnaG", "ftsA","hemD" )

data <- list(
  num_time = length(smooth_time),
  num_measurements = length(gse6865_raw_time),
  num_targets = length(targets),
  regulator_measured = 1,
  
  coeffs_prior_given = 0, #No coeffs prior. Sadly have to give at least size 1
  coeffs_prior_mean = array(0.0,1),
  coeffs_prior_cov = array(1, c(1,1)),
  
  measurement_times = measurement_times,
  num_spline_basis = num_spline_basis,
  spline_basis = spline_basis,
  expression = gse6865_raw[targets,],
  regulator_expression = gse6865_raw[source,],
  regulation_signs = array(1, length(targets)),
  measurement_sigma_absolute = 0.2,
  measurement_sigma_relative = 0.1,
  intercept_prior_sigma = 2,
  initial_condition_prior_sigma = 0.05,
  sensitivity_over_degradation_prior_mean = 2.5,
  sensitivity_over_degradation_prior_sigma = 2,
  sensitivity_prior_sigma = 1,
  degradation_prior_mean = -3,
  degradation_prior_sigma = 1,
  w_prior_sigma = 2,
  b_prior_sigma = 5,
  scale = 5
)

fit <- sampling(multiple_targets_model, data = data, control = list(adapt_delta = 0.95))
print(summary(fit, pars = c("coeffs", "intercept","initial_condition","sensitivity","degradation","w","b"))$summary)

num_samples <- 100
main_geom <- geom_line(alpha = 0.2)
samples_to_show <-  sample(1:4000, num_samples)
samples_regulator <- rstan::extract(fit,"predicted_regulator_expression")$predicted_regulator_expression[samples_to_show,]

regulator_plot <- ggmatplot(1:data$num_time, t(samples_regulator), main_geom = main_geom) + ggtitle(paste0("Regulator - ", source))

if(data$regulator_measured != 0) {
  regulator_plot <- regulator_plot + geom_point(data = data.frame(x = data$measurement_times,  y = data$regulator_expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)
}
print(regulator_plot)
 
samples_expression <- rstan::extract(fit,"predicted_expression")$predicted_expression[samples_to_show,,,drop=FALSE]
samples_expression_replicates <- rstan::extract(fit,"expression_replicates")$expression_replicates[samples_to_show,,,drop=FALSE]
for(target in 1:length(targets)) {
  print(ggmatplot(1:data$num_time, t(samples_expression[,target,]), main_geom = main_geom, x_title = "time", y_title = "expression") +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression[target,]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Expression - ", targets[target])))
  
  print(ggmatplot(data$measurement_times, t(samples_expression_replicates[,target,]), main_geom = main_geom, x_title = "time", y_title = "expression") +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression[target,]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Replicates - ", targets[target])))
  
}
```




##Checking whether resulting coeffs are approximately multivariate normal

```{r}
pairs(fit, pars = "coeffs")
samples_coeffs <- rstan::extract(fit,"coeffs")$coeffs
uniPlot(samples_coeffs, type ="qqplot")
uniPlot(samples_coeffs, type = "histogram")
mardiaTest(samples_coeffs, qqplot = TRUE)
hzTest(samples_coeffs)
n_coeffs <- dim(samples_coeffs)[2]
for(i in 1:(n_coeffs - 1)) {
  for(j in (i+1):n_coeffs) {
    test_res <- hzTest(samples_coeffs[,c(i,j)])
    mvnPlot(test_res, type = "contour")
  }
}
mvOutlier(samples_coeffs, qqplot = TRUE, method = "adj.quan")
```
We see there are some deviations from normality. In particular it seems the posterior has heavier tails than normal. Let's see what happens when we try to fit multivariete normal anyway. 

```{r}
coeffs_prior_cov <- cov(samples_coeffs)
chol_cov <- chol(coeffs_prior_cov)
coeffs_prior_mean <- colMeans(samples_coeffs)

means_array <- t(array(rep(coeffs_prior_mean, dim(samples_coeffs)[1]),c(dim(samples_coeffs)[2],dim(samples_coeffs)[1])))
samples_coeffs_mvn <- array(rnorm(dim(samples_coeffs)[1] * dim(samples_coeffs)[2], 0, 1), dim(samples_coeffs)) %*% chol_cov +  means_array

uniPlot(samples_coeffs_mvn, type = "histogram")
for(i in 1:(n_coeffs - 1)) {
  for(j in (i+1):n_coeffs) {
    test_res <- hzTest(samples_coeffs_mvn[,c(i,j)])
    mvnPlot(test_res, type = "contour")
  }
}

```


#Using the fitted normal to check regulations of other genes

```{r}
source <- "sigA"
#source <- "sigB"
#targets <- c("yfkJ")
targets <- c("yqxD","yfkJ","pyrB","purT")
#targets <- c("jag","kinC","kinE","lepA")


fits <- list()
for(t in 1:length(targets)) {
  cat("\n==== Target: ", targets[t],"\n")
  
  data <- list(
    num_time = length(smooth_time),
    num_measurements = length(gse6865_raw_time),
    num_targets = 1,
    regulator_measured = 0,
    
    coeffs_prior_given = 1, 
    coeffs_prior_mean = coeffs_prior_mean,
    coeffs_prior_cov = coeffs_prior_cov,
    
    measurement_times = measurement_times,
    num_spline_basis = num_spline_basis,
    spline_basis = spline_basis,

    expression = gse6865_raw[targets[t],,drop=FALSE],
    regulator_expression = numeric(0),
    regulation_signs = array(1, 1),
    measurement_sigma_absolute = 0.2,
    measurement_sigma_relative = 0.1,
    intercept_prior_sigma = 2,
    initial_condition_prior_sigma = 0.05,
    sensitivity_prior_sigma = 1,
    degradation_prior_mean = -3,
    degradation_prior_sigma = 1,
    w_prior_sigma = 2,
    b_prior_sigma = 5,
    scale = 5
  )
  
  fit <- sampling(multiple_targets_model, data = data, control = list(adapt_delta = 0.95))
  print(summary(fit, pars = c("coeffs", "intercept","initial_condition","sensitivity","degradation","w","b"))$summary)
  
  num_samples <- 100
  main_geom <- geom_line(alpha = 0.2)
  samples_to_show <-  sample(1:4000, num_samples)
  samples_regulator <- rstan::extract(fit,"predicted_regulator_expression")$predicted_regulator_expression[samples_to_show,]
  
  # regulator_plot <- ggmatplot(1:data$num_time, t(samples_regulator), main_geom = main_geom) + ggtitle(paste0("Regulator - ", targets[t])) + geom_point(data = data.frame(x = data$measurement_times,  y = gse6865_raw[source,]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)
  # print(regulator_plot)
  
  samples_expression <- rstan::extract(fit,"predicted_expression")$predicted_expression[samples_to_show,,,drop=FALSE]
  samples_expression_replicates <- rstan::extract(fit,"expression_replicates")$expression_replicates[samples_to_show,,,drop=FALSE]

    print(ggmatplot(1:data$num_time, t(samples_expression[,1,]), main_geom = main_geom, x_title = "time", y_title = "expression") +
      geom_point(data = data.frame(x = data$measurement_times,  y = data$expression[1,]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Expression - ", targets[t])))
    
    print(ggmatplot(data$measurement_times, t(samples_expression_replicates[,1,]), main_geom = main_geom, x_title = "time", y_title = "expression") +
      geom_point(data = data.frame(x = data$measurement_times,  y = data$expression[1,]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Replicates - ", targets[t])))
 
  fits[[t]] <- fit   
}
```
TODO: zkorelovat WAIC vs. sum sq.

```{r}

looic <- array(-1,length(targets))
waic <- array(-1,length(targets))
for(t in 1:length(targets)) {
  samples_log_lik <- rstan::extract(fits[[t]], "log_likelihood")$log_likelihood
  lik_data <- samples_log_lik[,1,]
  loo_res <- loo(lik_data)
  looic[t] <- loo_res$looic
  waic[t] <- waic(lik_data)$waic
}
o <- order(looic)
data.frame(order = (1:length(targets))[o],gene = targets[o], waic = waic[o], looic = looic[o])

```

##Comparing to constant synthesis
```{r}
csynth_model <- stan_model(file = 'constant_synthesis.stan')
```

```{r}
fits_csynth <- list()
for(t in 1:length(targets)) {
  
  data <- list(
    num_measurements = length(gse6865_raw_time),
    measurement_times = gse6865_raw_time + 1,
    expression = gse6865_raw[targets[t],],
    measurement_sigma_absolute = 0.2,
    measurement_sigma_relative = 0.1,
    initial_condition_prior_sigma = 0.05,
    synthesis_over_degradation_prior_mean = 0,
    synthesis_over_degradation_prior_sigma = 15,
    degradation_prior_mean = -3,
    degradation_prior_sigma = 1
  )  
  fit_csynth <- sampling(csynth_model, data = data, control = list(adapt_delta = 0.95))
   summary(fit_csynth)$summary[,c("n_eff","Rhat")] %>% as.data.frame() %>% summarise(min_n_eff = min(n_eff),max_Rhat = max(Rhat))
  
  num_samples <- 100
  main_geom <- geom_line(alpha = 0.2)
  samples_to_show <-  sample(1:4000, num_samples)
  
  
  samples_expression <- rstan::extract(fit_csynth,"predicted_expression")$predicted_expression[samples_to_show,]
  print(ggmatplot(data$measurement_times, t(samples_expression), main_geom = main_geom) + 
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(targets[t]))
  
  samples_replicates <- rstan::extract(fit_csynth,"expression_replicates")$expression_replicates[samples_to_show,]
  print(ggmatplot(data$measurement_times, t(samples_replicates), main_geom = main_geom) +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(targets[t]))
  
  fits_csynth[[t]] <- fit_csynth
}
```
```{r}

looic <- array(-1,length(targets))
waic <- array(-1,length(targets))
for(t in 1:length(targets)) {
  samples_log_lik <- rstan::extract(fits_csynth[[t]], "log_likelihood")$log_likelihood
  lik_data <- samples_log_lik
  loo_res <- loo(lik_data)
  looic[t] <- loo_res$looic
  waic[t] <- waic(lik_data)$waic
}
o <- order(looic)
data.frame(order = (1:length(targets))[o],gene = targets[o], waic = waic[o], looic = looic[o])

```


##Without known regulator

```{r}
source <- "sigA"
targets <- c("codY","clpQ","yqxD") #,"dnaA","yfkJ","yhdL", "pyrB","purT"

data <- list(
  num_time = ceiling(length(smooth_time) / 2),
  num_measurements = length(gse6865_raw_time),
  num_targets = length(targets),
  regulator_measured = 0,
  measurement_times = floor(gse6865_raw_time / 2) + 1,
  num_knots = num_knots,
  knots = knots / 2,
  spline_degree = 3,
  expression = gse6865_raw[targets,],
  regulator_expression = numeric(0),#gse6865_raw[source,],
  regulation_signs = array(1, length(targets)),
  measurement_sigma_absolute = 0.2,
  measurement_sigma_relative = 0.1,
  intercept_prior_sigma = 0.01,
  initial_condition_prior_sigma = 0.05,
  sensitivity_prior_sigma = 1,
  degradation_prior_mean = -3,
  degradation_prior_sigma = 1,
  w_prior_sigma = 2,
  b_prior_sigma = 5,
  scale = 5
)

fit <- sampling(multiple_targets_model, data = data, control = list(adapt_delta = 0.95))
print(summary(fit, pars = c("coeffs", "intercept","initial_condition","sensitivity","degradation","w","b"))$summary)

num_samples <- 100
main_geom <- geom_line(alpha = 0.2)
samples_to_show <-  sample(1:4000, num_samples)
samples_regulator <- rstan::extract(fit,"predicted_regulator_expression")$predicted_regulator_expression[samples_to_show,]

regulator_plot <- ggmatplot(1:data$num_time, t(samples_regulator), main_geom = main_geom) + ggtitle(paste0("Regulator - ", source))

if(data$regulator_measured != 0) {
  regulator_plot <- regulator_plot + geom_point(data = data.frame(x = data$measurement_times,  y = data$regulator_expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)
}
print(regulator_plot)
 
samples_expression <- rstan::extract(fit,"predicted_expression")$predicted_expression[samples_to_show,,,drop=FALSE]
for(target in 1:length(targets)) {
  print(ggmatplot(1:data$num_time, t(samples_expression[,target,]), main_geom = main_geom) +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression[target,]), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Expression - ", targets[target])))
}
  

```

#Older models

##Unknown regulator

```{r}
unknown_reg_model <- stan_model(file = 'unknown_regulator_splines.stan')
```

```{r}
fits <- list()
```

```{r}
#Genes fit by sigA that don't really have a sigA-like regulator profile:
# yfkJ, yhdL (requires smaller abs. error)
#Those where it works: codY, clpQ, purT, yqxD
#Some not regulated by sigA: pyrB, dnaA
targets <- c("codY","clpQ","purT","yqxD", "yfkJ","yhdL", "dnaA", "pyrB")
for(target in targets) {

  data <- list(
    num_time = length(smooth_time),
    num_measurements = length(gse6865_raw_time),
    measurement_times = gse6865_raw_time + 1,
    num_knots = num_knots,
    #knots = quantile(gse6865_raw_time + 1, probs = seq(1 / (num_knots + 1) ,1 - (1/ (num_knots + 1)), length.out = num_knots)),
    knots = knots,
    spline_degree = 3,
    expression = gse6865_raw[target,],
    measurement_sigma_absolute = 0.5,
    measurement_sigma_relative = 0.2,
    sensitivity_prior_sigma = 1,
    degradation_prior_sigma = 0.3,
    scale = 5
  )
  
  fit <- sampling(unknown_reg_model, data = data, control = list(adapt_delta = 0.95))
  cat("\n ======== ",target," =======\n")
  print(summary(fit, pars = c("coeffs","initial_condition","sensitivity","degradation", "intercept"))$summary)
  
  num_samples <- 100
  main_geom <- geom_line(alpha = 0.2)
  samples_to_show <-  sample(1:4000, num_samples)
  samples_regulator <- rstan::extract(fit,"regulator_profile")$regulator_profile[samples_to_show,]
  print(ggmatplot(1:data$num_time, t(samples_regulator), main_geom = main_geom) + ggtitle(paste0("Regulator - ", target)))
   
  samples_expression <- rstan::extract(fit,"predicted_expression")$predicted_expression[samples_to_show,]
  print(ggmatplot(1:data$num_time, t(samples_expression), main_geom = main_geom) +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Expression - ", target)))
  
  fits[[target]] <- fit
}
```

```{r}
possible_source <- "sigA"

for(target in targets) {
  corr_dist <- array(-100, 4000)
  samples_regulator <- rstan::extract(fits[[target]],"regulator_profile")$regulator_profile
  source_profile_splined <- gse6865_splined[possible_source,]
  for(s in 1:4000) {
    corr_dist[s] <- cor(source_profile_splined, samples_regulator[s,])
  }
  
  cat("\n ======== ",target," =======\n")
  print(ggplot(data.frame(corr = abs(corr_dist)), aes(x=corr)) + geom_histogram(bins = 30) + ggtitle(paste0("Abs. corr - ", target)))
  cat("P(|cor| > 0.9) = ", mean(abs(corr_dist) > 0.9), "\n")
  cat("#(|cor| > 0.9) = ", sum(abs(corr_dist) > 0.9), "\n")
  cat("E(|cor|) = ", mean(abs(corr_dist)), "\n")
  cat("E((1-|cor|)^2) = ", mean((1 - abs(corr_dist))^2), "\n")
#  mean(exp(-(1-abs(corr_dist))^2))
#  ggplot(data.frame(corr_exp = exp(-(1-abs(corr_dist))^2)), aes(x=corr_exp)) + geom_histogram(bins = 30)
}
```
```{r}
normal_reg_model <- stan_model(file = 'normal_regulation_splines.stan')
```


```{r}
source <- "sigA"
targets <- c("codY","clpQ","purT","yqxD", "yfkJ","yhdL", "dnaA", "pyrB")
for(target in targets) {

  data <- list(
    num_time = length(smooth_time),
    num_measurements = length(gse6865_raw_time),
    measurement_times = gse6865_raw_time + 1,
    num_knots = num_knots,
    #knots = quantile(gse6865_raw_time + 1, probs = seq(1 / (num_knots + 1) ,1 - (1/ (num_knots + 1)), length.out = num_knots)),
    knots = knots,
    spline_degree = 3,
    expression = gse6865_raw[target,],
    regulator_expression = gse6865_raw[source,],
    measurement_sigma_absolute = 0.5,
    measurement_sigma_relative = 0.2,
    initial_condition_prior_sigma = 0.05,
    sensitivity_prior_sigma = 1,
    degradation_prior_sigma = 0.3,
    w_prior_sigma = 2,
    b_prior_sigma = 5,
    scale = 5
  )
  
  fit <- sampling(normal_reg_model, data = data, control = list(adapt_delta = 0.95))
  cat("\n ======== ",target," =======\n")
  print(summary(fit, pars = c("coeffs", "intercept","initial_condition","sensitivity","degradation","w","b"))$summary)
  
  num_samples <- 100
  main_geom <- geom_line(alpha = 0.2)
  samples_to_show <-  sample(1:4000, num_samples)
  samples_regulator <- rstan::extract(fit,"predicted_regulator_expression")$predicted_regulator_expression[samples_to_show,]
  print(ggmatplot(1:data$num_time, t(samples_regulator), main_geom = main_geom) +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$regulator_expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3)+ ggtitle(paste0("Regulator - ", target)))
   
  samples_expression <- rstan::extract(fit,"predicted_expression")$predicted_expression[samples_to_show,]
  print(ggmatplot(1:data$num_time, t(samples_expression), main_geom = main_geom) +
    geom_point(data = data.frame(x = data$measurement_times,  y = data$expression), aes(x=x, y=y), inherit.aes = FALSE, color = "#ba1b1d", size = 3) + ggtitle(paste0("Expression - ", target)))
  
  fits[[target]] <- fit
}
```



